Visualizing Project Evolution Through Abstract Syntax Tree Analysis

### 2016-06-27 paper/document.tex
New intro.

Source code is often represented as text in a source code file. This text is then parsed by a compiler or interpreter with the intent of eventually executing the code represented by the source text. While programmers may type in individual characters, their intention is at a much higher level. Their intent is written in tokens that represent statements, identifiers, and types that are used to solve their problems. Programmers solve problems in a way that is readable to other developers and executable by a computer. The programmer rarely operates at the character level, or even the line level, so why should our visualizations of source code operate at granularities unfamiliar to developers?

The plain text nature of software does not stop at the source code level. The history and evolution of a software project is often tracked by version control systems, such as Git, in a programming language agnostic way. Updates to source code, documentation, and tests are recorded in the form of commits, which store the differences in lines of text in a file between each subsequent revision of the software. Version control systems become vital for large projects since individual changes would be impossible to track manually. Large projects have thousands of individual changes made by many different developers. With more changes to peruse, it becomes more difficult to see how a project is evolving over time since there is so much information to look through. Hosting services for Git repositories often provide some basic tools for analyzing projects over time~\cite{github-graphs,bitbucket-graphs}; however, as we will show, these tools often lack the kind of detail to gather insightful analysis. As with source code, commits operate at the text level. While commits may store individual changes as lines of code, they imply a change in the \emph{structure} of software.  By visualizing lines changed, or files changed, many current tools for software projects rarely capture the \emph{meaning} behind each commit; lines of code alone do not describe how a repository changed at a particular point in time. Thus, information regarding contributions from developers is often misleading since the number of commits and lines of code changed do not necessarily reflect meaningful changes to a code repository~\cite{robles2014}. In this paper, we will show that abstract syntax trees (ASTs) can be used to create \emph{semantically-aware} software evolution visualizations.

Consider GitHub graphs~\cite{github-graphs}: visualizations that are easily available to software developers using GitHub to host their Git software repositories. GitHub's graphs use line and commit information exclusively. In the case of the ANTLR4 project (Figure~\ref{fig:GH_code_freq}), this code frequency chart creates more questions than it answers. What happened at those big spikes of line changes in the graph? One of the spikes is so large, we had to crop the image in order to show the magnitude of the other changes. How did these large changes affect the code base? Who committed these large changes? Was it a team of contributors or one particularly prolific programmer? Plotting the number of lines changed may be misleading. A huge spike in additions may be caused by a number of reasons, such as the inclusion of automatically-generated documentation, the inclusion of an external library, or perhaps a massive refactoring effort~\cite{Hindle:2008:LCT:1370750.1370773}.

The goal of a software evolution visualization is to provide an alternate view that provides insight into the development of the software that is otherwise unavailable. By visualizing changes to lines and files, one is tracking the changes in text---not software. Instead, we promote \emph{semantically-aware visualizations} that visualize the structural changes in code. In our case these are changes in variable declarations, invocations of methods, and usage of types. Others have explored semantically-aware visualization~\cite{sourcevis,6980224,Kim:2009:DRS:1555001.1555046,7332410}, however, we show that this can be done with ASTs.

In this paper, we propose a semantically-aware visualization method that tracks the changes in type usage and method invocations over time. Using commits, we are able to keep track of \emph{who} is using what types. We then present this data using time-proven visualizations. We claim that type usage and method invocation offers more insight than line-based, file-based, and commit-based methods of visualization and analysis. For simplicity, we analyze only Java projects, as it allowed us to quickly develop and test our software. However, since the approach described in this paper can be generally applied to any statically-typed language, it can be adapted to other programming languages. Our goal is to describe a methodology for obtaining richer, more meaningful data from source code repositories which can be used for more insightful analyses.

In the following sections of the paper we will describe our tool, TypeV, that we developed to compute and visualize changes to types by author used in a project over time. TypeV is able to extract type information from a repository by computing commit-by-commit abstract syntax tree differences, which we describe in Section~\ref{sec:methodology}. In this paper, we answer the following questions: \\ \\
\textbf{RQ1:} How can we extract type and method usage from repositories over time? \\ \\
\textbf{RQ2:} How does type coverage compare to file coverage?
\\ \\
\textbf{RQ3:} Does the type information give any interesting or useful insight into the project?

@@

One issue with analyzing individual authors is that authors can commit from a variety of different computers which can have different emails and usernames attached to their Git configuration. This may result in one author appearing as several different authors in the \texttt{git log}. There is no perfect way to deal with this problem; however, we added an interface for managing authors. A user can manually specify an author's various aliases using a simple dropdown list. All statistics attributed to an author's various aliases are then remapped as belonging to one canonical author as specified by the user.

In addition, our author managing view also provides a facility for selecting and deselecting authors in the view. Thus, visualization users can focus on the changes made by one author or a group of authors, rather than only having the option of view changes made by all authors cumulatively.

>>>>

Visualizations can assist our understanding of a software repository. The history of a code base is often tracked in a version control system such as Git or SVN. Popular source code repository hosting websites such as GitHub~\cite{github-graphs} and BitBucket~\cite{bitbucket-graphs} often provide a way to visualize project evolution. These visualizations often show changes to lines of code, as line changes are what the underlying version control system tracks. The advantage of visualizing line changes is that it works for any text-based file. There is a major disadvantage, however: by only visualizing changes to lines of code, GitHub (Figure~\ref{fig:GH_code_freq}) and similar visualizations lose the \emph{meaning} behind those changes. Although a programmer may type individual characters into their text editor, the lines of code represent something much deeper. A Java programmer, for instance, writes lines of code with the \emph{intent} of creating packages, defining classes, declaring fields, and implementing methods. An author commits lines of code to implement features, write tests, fix bugs, and refactor code bases. Lines of code merely tell us that changes were made; not the \emph{reason} for the changes.

To address the lack of semantically-aware visualizations we created \textbf{TypeV}. Instead of visualization changes to lines of code, we visualize changes to type usages and invocations. By using types instead of lines of code, our visualization can better answer the high-level questions programmers often ask~\cite{sillito2006}. TypeV is composed of two parts: a data extraction application that performs commit-by-commit AST differencing to extract type changes; and an interactive web visualization that allows for the exploration of a project's evolution. Using TypeV, we seek to answer the following research questions:
  \\ \\
\textbf{RQ1:} How can we extract type and method usage from repositories over time?
  \\ \\
\textbf{RQ2:} How does type coverage compare to file coverage?
  \\ \\
\textbf{RQ3:} What extra insight can we gain from type-based visualization?

@@

One issue with analyzing source code repositories is that authors can commit from a variety of different computers which can have different emails and usernames attached to their Git configuration. This may result in one author appearing as several different authors in the \texttt{git log}. There is no perfect way to deal with this problem; however, we added an interface for managing authors. A user can manually specify an author's various aliases using a simple dropdown list. All statistics attributed to an author's various aliases are then remapped as belonging to one canonical author as specified by the user. The author managing view also provides a facility for selecting and deselecting authors shown in the timeline. Thus, visualization users can focus on the changes made by a particular author or a group of authors, rather than having to view the changes made by all authors.
